Simplex: An individual-based platform for ecological modeling

### 2015-11-03 paper/PrePrint.Rmd
edited PrePrint.pdf and .Rmd -KL
edited PrePrint.Rmd -KL

---
title: "simplex: An individual-based platform for ecological modeling"
author: "Kenneth J. Locey"
fontsize: 12pt
geometry: margin=2.54cm
output: pdf_document
header-includes:
    - \usepackage{setspace}
    - \doublespacing
---

@@

A popular approach in ecological modeling explicitly simulates the biology and ecology of individual organisms. Individual-based models (IBM) allow ecological patterns and dynamics to be studied as they emerge from the individual level according to simulated principles and theories. IBMs also allow virtually any aspect of a simulated system to be tracked and recorded for later analysis. I developed a novel IBM platform (i.e, simplex) to explicitly simulate physiology, life history, resource-limited growth, fluctuating environmental dynamics, and community assembly and structure in both fluid and non-fluid environments. simplex accomplishes three primary tasks in keeping with its namesake (i.e., a generalized notion of a triangle to arbitrary dimensions). First, simplex assembles and runs stochastic ecological models from random combinations of over 20 state variables. Second, simplex stores the output of models and generates and stores animations. Third, simplex provides R Markdown files for analyzing simulated data. The simplex source code is freely available, includes the standard ODD description protocol, is open to community-based revision, includes unit testing and time testing, and has been developed to run on modest hardware.

@@

Modeling is a elementary approach to understanding ecological systems, the influence of ecological processes, and the predictability of ecological patterns and dynamics. Before modern computing, ecological models were almost exclusively equation-based representations of highly simplified systems (Black and McKane 2012, Otto and Troy 2007, Grant and Swannack 2008). Since the advent of personal computing, ecological models have been increasingly constructed to handle greater complexity and to explicitly simulate ecological processes.

In ecology, simulation-based models are often used to examine analytically challenging or intractable scenarios. 
Examples are markov models that simulate stochastic demographic changes and early ecological null models that , both of which operate on some form of matrix (Gotelli and Entsminger 2001, Hubbell 2001). Ecologists have also simulated growth and interactions among individuals for over two decades with individual-based models (IBMs) (DeAngelis and Gross 1992, Rosindell et al. 2015). 
IBMs explicitly encode rules of how individuals change and interact. Once the rules are encoded, population to ecosystem-level dynamics can emerge as an IBM simulates over time and spatially explicit environments.

@@

Despite their frequent use, the power of ecological IBMs has yet been leveraged to its greatest advantage. Few, if any, include fluid dynamics or combine fluid dynamics with growth and dispersal. While growth, sensing, and decision making are often modeled, few IBMs integrate physiology, nutrient-limited growth, evolution, environmental flux, community ecology, biogeography, and stochastic sampling theory. Ecological IBMs are more often constructed to model a specific system than to synthesize general theories of ecology and evolution (but see Rosindell et al. 2015). Yet, IBMs can allow researchers to creatively and explicitly simulate any number of biological and ecological theories and principles. Likewise, IBMs can made to track and analyze any information from physiological changes to community assembly, and the structure of an entire ecosystem.

Here, I present a detailed description for an ecological IBM platform referred to as "simplex". The simplex platform simulates a broad range of ecological conditions, records information from individuals to the ecosystem level, and provides computing code for analyzing output. simplex incorporates fluid dynamics, stochastic and selection-based changes within and among populations, and integrates nutrient-limited growth and physiology. I constructed simplex with the aim of simulating ecologically complex scenarios, from which the results could be used to test (*i*) predictions of ecological theory and (*ii*) the general influence of constraints and processes on populations, communities, and ecosystems. I provide an explanation of how simplex works, the data it quantifies and tracks, the theories and principles that simplex integrates, and the analyses that can be conducted using the open source code.

@@

**Model description following ODD Protocol**  
The ODD protocol is standard for describing individual-based models (Grimm et al. 2006). ODD stands for Overview, Design concepts, and Details. Here, I provide an abbreviated explanation of the simplex ODD protocol. The full ODD protocol can be found on the public GitHub repository at https://github.com/LennonLab/simplex. Likewise, to save space, I refer the reader to the README.md file on the public simplex GitHub.com repository (https://github.com/LennonLab/simplex/blob/master/README.md) for descriptions of source code files and directories.

**Purpose**  
The purpose of simplex is to explicitly simulate life history and the evolution of traits among individual organisms, the assembly of ecological communities, and the structure and function of ecosystems in a spatially explicit environment under stochastic conditions. The proximate goal of simplex is to generate high degrees of variation in the assembly and structure of populations and communities by assembling many different models from random combinations of state-variables and processes. The ultimate goal of simplex is to provide a simulation-based platform for examining conditions under which processes and constraints have a robust influence on eco-evolutionary dynamics, patterns of biodiversity, and ecosystem processes.

@@

*Individual organisms*--Individuals are distinguished by collections of elements within lists. Individuals undergo changes when randomly sampled from lists. Each specific position in the list corresponds to the same individual. For example, in simulating growth, a simplex model chooses an individual from the list of cell quotas (the probability of reproducing is determined by endogenous resources). The first position in this list as well in all other lists of individual attributes corresponds to the same individual. Lists of attributes for individual organisms include the time each individual spends in the system, species ID, individual ID, 2D spatial location, endogenous levels of resources (resource specific cell quotas), individual-level metabolic maintenance cost, individual-level maximum dispersal rate, parental lineage, individual resource use efficiency for each resource.

*Species*--Each species is characterized by the individuals that share a common set of traits, such as maximum growth rate, metabolic maintenance cost. Species information is stored in Python dictionaries. In this way, if simplex requires the species ID of an individual it will access the species ID list where each element corresponds to a specific individual. But, if simplex requires the maximum growth rate for an individual, then it finds the species ID and then uses that to accesses the Python dictionary for maximum specific growth rates of species. The species-level information that are stored include metabolic maintenance cost, maximum dispersal rate, maximum theoretical growth rate, and the resource use efficiency for each resource.

@@

*Inert tracers particles*--These are objects that move/flow into and through the environment. They only interact with physical barriers. The use of tracers particles allows for attributes of a system that flows or physically turns over to be quantified. Like other individual-level objects in simplex models, tracers are distinguished by collections of elements within lists. Each specific position in the list corresponds to the same resource particle. Information stored about each resource particle includes time spent in the system, 2D spatial location and particle ID.

*Physcial barriers*--These objects are simulated as discrete 2D spatial coordinates that cannot be occupied by any individual entities. The number, size, and location of physical barriers are chosen at random by simplex at the start of each model.

**System level state variables**  
Each run of simplex begins with random choices for the values of width (5 to 100), height (5 to 100), basal flow rate (0.0 to 1.0), the number, size, and location of physical barriers, the number and direction of environmental gradients, rate of stochastic disturbance, rate of fluctuation in basal flow rate, and degrees of fluctuation and syncronized flow of individual particles, ranging from completely in-sync or completely out of sync.

**Spatial and temporal scales**  
\setlength\parindent{36pt}
The two general aspects of scale are grain (i.e., resolution) and extent (e.g. total area). The environment of simplex models is two dimensional and can vary along each axis from 5 to 100 discrete units. This makes for a potential total extent of 25 to 10,000 discrete patches, each with a grain of 1 square unit. All particles move in decimal units, meaning that individual particles can  practically infinite locations within patches and likewise. Extent of time in simplex models refers to residence time, i.e., the average amount of time that individual particles spend in the system. Residence time for individual particles (organisms, resources, tracers) can vary across five orders of magnitude.

Grain is the smallest unit over which change can happen. For example, the original ODD documentation states “One time step represents one year and simulations were run for 100 years. One grid cell represents 1 ha and the model landscape comprised 1,000 x 1,000 ha; i.e., 10,000 square kilometers”. In contrast, the smallest grain achievable by simplex is determined by slowest rate at which individuals can undergo BIDE (birth, immigration, death, emigration) processes. For example, under high residence times, an individual can move across 0.00001% of the x or y axis in one time step. Under low residence times, an individual can move across 10% or greater of the x or y axis in one time step.

@@

*Core simulation process*. simplex models begin simulation immediately after assembly from random combinations of state-variables and processes. Instead of operating by definitive time steps (i.e. days, generations), simplex models advance turnover of the environmental matrix according to the initial rate of flow. If the initial rate of flow is 1.0, then the environmental matrix and inert particles would flow 1.0 units of distance. After each iteration of flow, each individual is given the chance to consume, grow, reproduce, starve, die, and to disperse towards resources and environmental optima.

*Duration: A run to mean reversion*. Once assembled, a simplex model simulates ecological processes (birth, death, dispersal, growth, consumption, etc.) until the system reaches a point of mean reversion, i.e., the tendency of a system to reverse a directional change in, say, total abundance. Mean reversion captures whether a system is fluctuating around a mean value. Once 100 generations have occurred within a model, simplex examines whether a point of mean reversion has occured by conducting an Augmented Dickey-Fuller (ADF) Test, which is well-explained here: https://www.quantstart.com/articles/Basics-of-Statistical-Mean-Reversion-Testing. A model is stopped once mean reversion is determined to occurred and, unless the number of desired simulations (i.e. models) has been reached, simplex simply constructs another model from random combinations of state-variables and processes, and then runs it to mean reversion.

*Fluid dynamics*. simplex uses an efficient and powerful method for simulating fluid flow, i.e., a Lattice-Boltzmann Method (LBM) (Succi 2001). An LBM discretizes the environment into a lattice and attaches to each position in the lattice a number of particle densities and velocities for each simulated direction. simplex simulates nine directions of movement possible in a 2D spatial environment (i.e., N, S, E, W, NE, NW, SE, SW, current position).

*Active dispersal*. simplex models allow individuals to move towards their environmental optima. Rather than a single environmental optima resulting from a single environmental gradient, simplex allows environmental optima to occur as intersections among environmental gradients. Hence, individuals potentially have multiple optima resulting from unique and equally optimal intersection of up to 10 environmental gradients.

*Simulated life history*.\setlength\parindent{36pt} simplex models simulate growth, reproduction, and death via weighted random sampling. This simulate the partly probabilistic and partly deterministic nature of environmental filtering and individual-level interactions.

@@

Reproduction. Reproduction in simplex is currently limited to clonal reproduction with the possibility of mutation. Individuals reproduce with a probability determined by the combination of mean cell quota and the proportional match to the environmental optima. The cell quota of each resource is evenly divided between the individual and its daughter. The daughter is given a unique individual ID and the species ID of its mother, unless in the case of speciation, but is allowed small mutations in individual-level state variables.

Speciation. Speciation is simulated within simplex as a discrete event (*sensu* Hubbell 2001 and Bell 2001) and is accompanied by mutations in the values of species-level state variables, much like that expected from a Wright-Fisher model (Hartl and Clark 1997). These random events allows for diversity to arise within the system, which the environmental filter can then select on.

@@

**Design concepts**  
*Ecological complexity*. simplex assembles models from random combinations of constraints (state-variables) and processes to generate output data that allow the user to test the general influence of a particular state-variable, process, or combination thereof using univariate and multivariate tests.

*Nutrient limited growth*. All models assembled by simplex employ the universal concept that individual growth and activity is fueled and limited by resources.

*Resource diversity & heterogeneity*. simplex allows the user to explore the influence of the number and abundances of different resources on ecological diversity and ecosystem processes.

**Theories**  
*Constraint-based theory*. simplex was originally built to explore the influence of ecosystem residence time (volume/flow rate) on community assembly and structure. That is, the idea that both ecological processes and physical constraints shape ecological diversity.

*Chemostat theory*. simplex operates much like an unhinged bioreactor or chemostat. That is, particles flow through a system of a defined size at an average rate, and are limited in growth by their residence time.

*Ecological neutral theory*. simplex operates via random sampling and can vary from being completely neutral (all individuals having equal vital rates) to completely idiosyncratic (all individual and species are as different as possible). The one aspect of neutral theory that simplex adopts without question is the importance of stochastic life history processes (i.e. weighted or unweighted random fluctuations in population sizes).

@@

**Observation**  
The following is recorded for each simplex model:

@@

**Initialization**  
The model initiates with a random set of values for state-variables, 100 to 10,000 randomly drawn individuals from a theoretical log-series metacommunity. These values are saved, so that a simplex model could be programmed to replicate an analysis.

**Input data**  
simplex models require no input data, but it might be cool to use an api to grab environmental data or other 
data from a website to parameterize a simplex model.

@@

*Cell quota model of Droop*. In simplex models, individuals grow according to their amounts of endogenous resources (cell quota).

@@

*Maintenance cost of Pirt*. Pirt (1965) states "The variation, with growth rate, of the yield of organism from the substrate used as energy source is attributed to consumption of energy at a constant rate for cell maintenance." He derived a relationship between the growth yield (biomass), the growth rate, and metabolic maintenance. simplex models use Pirt's concept of a constant maintenance requirement. simplex also draws from Pirt's simple relation for substrate use: $$use(total) = use(maintenance) + use(growth)$$

@@

*Log-series metacommunity*. simplex models draw immigrating individuals from a theoretical log-series distribution.
Hubbell (2001) states that the regional community (i.e., metacommmunity) often has a log-series species 
abundance distribution. The probability density function (pdf) of the log-series is: $$f(k) = -1/ln(1 - p) * p^{k}/k$$

**Notes on simplex source code**  
\setlength\parindent{36pt}
simplex models operate primarily on lists in a programmatic way, e.g., quickly sorting lists, and removing and returning an element from lists with very little overhead. Likewise, simplex models generate and hold a lot of information about all the particles and elements in the system, which can become a computationally intensive task. To this end, simplex modeling coded is written in Python, an easy to read high-level programming language that has many scientific, plotting, and animation libraries. Python gives greater control over the operating system than data analysis languages (e.g. R, Matlab) that can be comparatively slow at purely computational tasks and can greatly limit the amount of memory held in any data object, and even fail to import large amounts of data. Python can also obtain C-like speeds when implementing certain software, e.g., Cython, Sage.

The output of simplex is a broad array of information held in seven .csv files. The most important of these is SimData.csv, and is intended to be analyzed in the freely available R (https://www.r-project.org/) and RStudio (https://www.rstudio.com/) environments. The R statistical computing language is well-suited to the analysis of simplex output and contains many packages for multivariate analysis and higher-order statistical analysis that Python is only beginning to accumulate. Consequently, I provide R source code in .R files and .Rmd (RMarkdown) files, complete with basic and advanced statistical analyses for analyzing diversity, regression models, ordination, variance partitioning, and for generating pdf documents (via Knitr) that integrate prose, code, and figures for manuscripts.

The reader can view example R-based analysis files that users can use to examine simplex's simulated data: https://github.com/LennonLab/simplex/tree/master/results/analyses.

@@

\setlength\parindent{36pt}
Simplex models do not complete until the time series of total abundance values reaches a state of mean reversion (i.e., stationarity). Because simplex models can range from quickly flowing systems with high disturbance to barely flowing and highly stable but depleted systems, simulations can potentially take several minutes or more to complete. Likewise, the ability to simulate many complex scenarios also allows for very large total abundances, the values of which cannot be predicted *a priori* and can even potentially outstrip a computer's memory.

I ran simplex on a Mid 2010 MacBook Pro (OS X 10.9.5) with a 2.4 GHz Intel Core 2 Duo processor and 4GB of Memory. This system probably represents a below average capacity for modern personal computers, which for this study, was desirable as simplex should be able to be ran on both personal computers and high capacity remote servers. I present results for time to completion and required memory in the Results.

@@

**Unit tests**  
simplex passed all units tests for 15 diversity indices, ensuring that each index returns either the correct calculated value, or 'NaN' if given any values that cannot be used (e.g., negative numbers, string characters, empty lists).

**Speed & Memory**  
The following are results from 100 randomly assembled models. On average, simplex models required 96.60 +- 32.70 seconds to run, had an average total abundance (*N*) of 31,367 +- 879, and required 161 +- 4.3 megabytes of memory. The longest any model took to complete was 49 minutes. This particularly demanding model required 261MB of memory and also had generated the greatest total abundance (*N* = 68,808). The shortest any model took to complete was 5.8 seconds with a total abundance of 0 individuals. The least amount of memory any model required was 90MB. These and other analyses from the 100 randomly assembled models can be run using the SpeedMemory.Rmd file located in the results/TestResults directory.

@@

There are currently two R Markdown analysis files associated with simplex. These have been included as supplemental files. These are located in the tools/analyses/Rmd directory and are designed to help users understand how to import simplex files and use R Markdown to analyze simplex data. The first file, BasicAnalysis.Rmd, focuses on descriptive statistics, multiple univariate analyses, and variance partitioning to reveal relationships among simplex variables. Running this file on example simplex data revealed that simplex produces well-known and/or intuitive relationships (Figs 1, 2, 3, 4).

@@

**Figure 1.** A scatter plot matrix revealing that the values of inputs representing physical variables are independent of one another, as should be expected in the modeling. This is only a subset of the physical variables in simplex. Amplitude pertains to magnitude of temporal fluctuation in the inflow of propagules and/or resources. Barriers pertains to the number of barriers in the system, which varied between 1 and 4.

@@

**Figure 2.** A scatter plot matrix revealing that the values of resource-related variables. This is only a subset of resource variables in simplex. Number of resource particles, their concentration, resource richness, and resource diversity (a combination of resource richness and the variance in abundance among resources) are all highly and positively correlated, as should generally be expected.

@@

**Figure 3.** A scatter plot matrix revealing that the values of physiological variables. This is only a subset of those in simplex. As of 29 October 2015, simplex does not include explicit stoichiometry. An intuitive life history trade-off arises in that increasing growth rate generally leads to increasing metabolic maintenance.

@@

**Figure 4.** A scatter plot matrix revealing that the values of diversity-related variables. This is only a subset of those in simplex. There are strong positive relationships between total numbers of individuals (total abundance) and total numbers of species (species richness) and between total abundance and the abundance of the most abundant species (Nmax). Each of these relationships is widely observed in ecological communities (Locey and Lennon 2015). Likewise, different measures of species turnover largely reflect one another, as would generally be expected.

@@

**Animations**   
I used the animation module in the matplotlib plotting library to generate avi formatted movies of simplex models. These can be viewed in any internet brower (https://github.com/LennonLab/simplex/tree/master/results/movies/examples). Currently, generating movies requires following three easy steps explained in model/model.py.

>>>>

---
title: "Simplex: An individual-based platform for ecological modeling"
author: "Kenneth J. Locey"
fontsize: 12pt
geometry: margin=2.54cm
output: pdf_document
header-includes:
    - \usepackage{setspace}
    - \doublespacing
---

@@

A popular approach in ecological modeling explicitly simulates the biology and ecology of individual organisms. Individual-based models (IBM) allow ecological patterns and dynamics to be studied as they emerge from the individual level according to simulated principles and theories. IBMs also allow virtually any aspect of a simulated system to be tracked and recorded for later analysis. I developed a novel IBM platform (i.e., Simplex) to explicitly simulate physiology, life history, resource-limited growth, fluctuating environmental conditions, and community assembly and structure in both fluid and non-fluid environments. Simplex accomplishes three primary tasks in keeping with its namesake (i.e., a generalized notion of a triangle to arbitrary dimensions). First, Simplex assembles and runs stochastic ecological models from random combinations of over 20 state variables. Second, Simplex stores the output of simulations, and generates and stores animations. Third, Simplex provides R Markdown files for analyzing simulation data. The Simplex source code is freely available, includes a standard ODD description protocol, is open to community-based revision, includes unit testing and time testing, and has been developed to run on modest hardware.

@@

Modeling is a elementary approach to understanding ecological systems, the influence of ecological processes, and the predictability of ecological patterns and dynamics. Before modern computing, ecological models were almost exclusively equation-based representations of highly simplified systems (Black and McKane 2012, Otto and Troy 2007, Grant and Swannack 2008). Since the advent of personal computing, ecological models have been increasingly constructed to handle greater complexity and to explicitly simulate ecological processes, forces, and mechanisms.

Simulation-based models are often used to examine analytically challenging or intractable scenarios. 
Examples are markov models that simulate stochastic demographic changes and early ecological null models, both of which operate on matrices (Gotelli and Entsminger 2001, Hubbell 2001). Ecologists have also simulated growth and interactions among individual organisms for over two decades with individual-based models (IBMs) (DeAngelis and Gross 1992, Rosindell et al. 2015). 
IBMs explicitly encode rules of how individuals change and interact. Once these rules are encoded, population to ecosystem-level dynamics can emerge as an IBM simulates over time and spatially explicit environments.

@@

Despite their frequent use, the power of ecological IBMs has yet been leveraged to its greatest advantage. Few, if any, include fluid dynamics or combine fluid dynamics with growth and dispersal. While growth, sensing, and decision making are often modeled, few IBMs integrate physiology, nutrient-limited growth, evolution, environmental flux, community ecology, biogeography, and stochastic sampling theory. Instead, ecological IBMs are more often constructed to model a specific system than to synthesize general theories of ecology and evolution (but see Rosindell et al. 2015). Yet, IBMs can allow researchers to creatively and explicitly simulate any number of theories and principles. Likewise, IBMs can made to track and analyze any information from physiological changes and population demography, to community assembly and the structure of entire ecosystems.

Here, I present a detailed description for an ecological IBM platform I refer to as "Simplex". The Simplex platform simulates a broad range of ecological conditions, records information from individuals to the ecosystem level, and provides computing code for analyzing output. Simplex incorporates fluid and non-fluid dynamics with stochastic and selection-based changes within and among populations. Simplex also integrates nutrient-limited growth and physiology. I constructed Simplex with the aim of simulating ecologically complex scenarios, from which the results could be used to test (*i*) predictions of ecological theory and (*ii*) the general influence of constraints and processes on populations, communities, and ecosystems. I provide an explanation of how Simplex works, the data it quantifies and tracks, the theories and principles that Simplex integrates, and the analyses that can be conducted using the open source code.

@@

**Model description following ODD Protocol**  
The ODD protocol is standard for describing individual-based models (Grimm et al. 2006). ODD stands for Overview, Design concepts, and Details. Here, I provide an abbreviated explanation of the Simplex ODD protocol. The full ODD protocol can be found on the public GitHub repository at https://github.com/LennonLab/simplex. Likewise, to save space, I refer the reader to the README.md file on the public Simplex GitHub.com repository (https://github.com/LennonLab/simplex/blob/master/README.md) for descriptions of source code files and directories.

**Purpose**  
The purpose of Simplex is to explicitly simulate life history and the evolution of traits among individual organisms, the assembly of ecological communities, and the structure and function of ecosystems in a spatially explicit environment under stochastic conditions. The proximate goal of Simplex is to generate high degrees of variation in the assembly and structure of populations and communities by assembling many different models from random combinations of state-variables and processes. The ultimate goal of Simplex is to provide a simulation-based platform for examining conditions under which processes and constraints have a robust influence on eco-evolutionary dynamics, patterns of biodiversity, and ecosystem processes.

@@

*Individual organisms*--Individuals are distinguished by collections of elements within lists. Individuals undergo changes when randomly sampled from lists. Each specific position in the list corresponds to the same individual. For example, in simulating growth, a Simplex model chooses an individual from the list of cell quotas (the probability of reproducing is determined by endogenous resources). The first position in this list as well in all other lists of individual attributes corresponds to the same individual. Lists of attributes for individual organisms include the time each individual spends in the system, species ID, individual ID, 2D spatial location, endogenous levels of resources (resource specific cell quotas), individual-level metabolic maintenance cost, individual-level maximum dispersal rate, parental lineage, individual resource use efficiency for each resource.

*Species*--Each species is characterized by the individuals that share a common set of traits, such as maximum growth rate, metabolic maintenance cost. Species information is stored in Python dictionaries. In this way, if Simplex requires the species ID of an individual it will access the species ID list where each element corresponds to a specific individual. But, if Simplex requires the maximum growth rate for an individual, then it finds the species ID and then uses that to accesses the Python dictionary for maximum specific growth rates of species. The species-level information that are stored include metabolic maintenance cost, maximum dispersal rate, maximum theoretical growth rate, and the resource use efficiency for each resource.

@@

*Inert tracers particles*--These are objects that move/flow into and through the environment. They only interact with physical barriers. The use of tracers particles allows for attributes of a system that flows or physically turns over to be quantified. Like other individual-level objects in Simplex models, tracers are distinguished by collections of elements within lists. Each specific position in the list corresponds to the same resource particle. Information stored about each resource particle includes time spent in the system, 2D spatial location and particle ID.

*Physcial barriers*--These objects are simulated as discrete 2D spatial coordinates that cannot be occupied by any individual entities. The number, size, and location of physical barriers are chosen at random by Simplex at the start of each model.

**System level state variables**  
Each run of Simplex begins with random choices for the values of width (5 to 100), height (5 to 100), basal flow rate (0.0 to 1.0), the number, size, and location of physical barriers, the number and direction of environmental gradients, rate of stochastic disturbance, rate of fluctuation in basal flow rate, and degrees of fluctuation and syncronized flow of individual particles, ranging from completely in-sync or completely out of sync.

**Spatial and temporal scales**  
\setlength\parindent{36pt}
The two general aspects of scale are grain (i.e., resolution) and extent (e.g. total area). The environment of Simplex models is two dimensional and can vary along each axis from 5 to 100 discrete units. This makes for a potential total extent of 25 to 10,000 discrete patches, each with a grain of 1 square unit. All particles move in decimal units, meaning that individual particles can  practically infinite locations within patches and likewise. Extent of time in Simplex models refers to residence time, i.e., the average amount of time that individual particles spend in the system. Residence time for individual particles (organisms, resources, tracers) can vary across five orders of magnitude.

Grain is the smallest unit over which change can happen. For example, the original ODD documentation states “One time step represents one year and simulations were run for 100 years. One grid cell represents 1 ha and the model landscape comprised 1,000 x 1,000 ha; i.e., 10,000 square kilometers”. In contrast, the smallest grain achievable by Simplex is determined by slowest rate at which individuals can undergo BIDE (birth, immigration, death, emigration) processes. For example, under high residence times, an individual can move across 0.00001% of the x or y axis in one time step. Under low residence times, an individual can move across 10% or greater of the x or y axis in one time step.

@@

*Core simulation process*. Simplex models begin simulation immediately after assembly from random combinations of state-variables and processes. Instead of operating by definitive time steps (i.e. days, generations), Simplex models advance turnover of the environmental matrix according to the initial rate of flow. If the initial rate of flow is 1.0, then the environmental matrix and inert particles would flow 1.0 units of distance. After each iteration of flow, each individual is given the chance to consume, grow, reproduce, starve, die, and to disperse towards resources and environmental optima.

*Duration: A run to mean reversion*. Once assembled, a Simplex model simulates ecological processes (birth, death, dispersal, growth, consumption, etc.) until the system reaches a point of mean reversion, i.e., the tendency of a system to reverse a directional change in, say, total abundance. Mean reversion captures whether a system is fluctuating around a mean value. Once 100 generations have occurred within a model, Simplex examines whether a point of mean reversion has occured by conducting an Augmented Dickey-Fuller (ADF) Test, which is well-explained here: https://www.quantstart.com/articles/Basics-of-Statistical-Mean-Reversion-Testing. A model is stopped once mean reversion is determined to occurred and, unless the number of desired simulations (i.e. models) has been reached, Simplex simply constructs another model from random combinations of state-variables and processes, and then runs it to mean reversion.

*Fluid dynamics*. Simplex uses an efficient and powerful method for simulating fluid flow, i.e., a Lattice-Boltzmann Method (LBM) (Succi 2001). An LBM discretizes the environment into a lattice and attaches to each position in the lattice a number of particle densities and velocities for each simulated direction. Simplex simulates nine directions of movement possible in a 2D spatial environment (i.e., N, S, E, W, NE, NW, SE, SW, current position).

*Active dispersal*. Simplex models allow individuals to move towards their environmental optima. Rather than a single environmental optima resulting from a single environmental gradient, Simplex allows environmental optima to occur as intersections among environmental gradients. Hence, individuals potentially have multiple optima resulting from unique and equally optimal intersection of up to 10 environmental gradients.

*Simulated life history*.\setlength\parindent{36pt} Simplex models simulate growth, reproduction, and death via weighted random sampling. This simulate the partly probabilistic and partly deterministic nature of environmental filtering and individual-level interactions.

@@

Reproduction. Reproduction in Simplex is currently limited to clonal reproduction with the possibility of mutation. Individuals reproduce with a probability determined by the combination of mean cell quota and the proportional match to the environmental optima. The cell quota of each resource is evenly divided between the individual and its daughter. The daughter is given a unique individual ID and the species ID of its mother, unless in the case of speciation, but is allowed small mutations in individual-level state variables.

Speciation. Speciation is simulated within Simplex as a discrete event (*sensu* Hubbell 2001 and Bell 2001) and is accompanied by mutations in the values of species-level state variables, much like that expected from a Wright-Fisher model (Hartl and Clark 1997). These random events allows for diversity to arise within the system, which the environmental filter can then select on.

@@

**Design concepts**  
*Ecological complexity*. Simplex assembles models from random combinations of constraints (state-variables) and processes to generate output data that allow the user to test the general influence of a particular state-variable, process, or combination thereof using univariate and multivariate tests.

*Nutrient limited growth*. All models assembled by Simplex employ the universal concept that individual growth and activity is fueled and limited by resources.

*Resource diversity & heterogeneity*. Simplex allows the user to explore the influence of the number and abundances of different resources on ecological diversity and ecosystem processes.

**Theories**  
*Constraint-based theory*. Simplex was originally built to explore the influence of ecosystem residence time (volume/flow rate) on community assembly and structure. That is, the idea that both ecological processes and physical constraints shape ecological diversity.

*Chemostat theory*. Simplex operates much like an unhinged bioreactor or chemostat. That is, particles flow through a system of a defined size at an average rate, and are limited in growth by their residence time.

*Ecological neutral theory*. Simplex operates via random sampling and can vary from being completely neutral (all individuals having equal vital rates) to completely idiosyncratic (all individual and species are as different as possible). The one aspect of neutral theory that Simplex adopts without question is the importance of stochastic life history processes (i.e. weighted or unweighted random fluctuations in population sizes).

@@

**Observation**  
The following is recorded for each Simplex model:

@@

**Initialization**  
The model initiates with a random set of values for state-variables, 100 to 10,000 randomly drawn individuals from a theoretical log-series metacommunity. These values are saved, so that a Simplex model could be programmed to replicate an analysis.

**Input data**  
Simplex models require no input data, but it might be cool to use an api to grab environmental data or other 
data from a website to parameterize a Simplex model.

@@

*Cell quota model of Droop*. In Simplex models, individuals grow according to their amounts of endogenous resources (cell quota).

@@

*Maintenance cost of Pirt*. Pirt (1965) states "The variation, with growth rate, of the yield of organism from the substrate used as energy source is attributed to consumption of energy at a constant rate for cell maintenance." He derived a relationship between the growth yield (biomass), the growth rate, and metabolic maintenance. Simplex models use Pirt's concept of a constant maintenance requirement. Simplex also draws from Pirt's simple relation for substrate use: $$use(total) = use(maintenance) + use(growth)$$

@@

*Log-series metacommunity*. Simplex models draw immigrating individuals from a theoretical log-series distribution.
Hubbell (2001) states that the regional community (i.e., metacommmunity) often has a log-series species 
abundance distribution. The probability density function (pdf) of the log-series is: $$f(k) = -1/ln(1 - p) * p^{k}/k$$

**Notes on Simplex source code**  
\setlength\parindent{36pt}
Simplex models operate primarily on lists in a programmatic way, e.g., quickly sorting lists, and removing and returning an element from lists with very little overhead. Likewise, Simplex models generate and hold a lot of information about all the particles and elements in the system, which can become a computationally intensive task. To this end, Simplex modeling coded is written in Python, an easy to read high-level programming language that has many scientific, plotting, and animation libraries. Python gives greater control over the operating system than data analysis languages (e.g. R, Matlab) that can be comparatively slow at purely computational tasks and can greatly limit the amount of memory held in any data object, and even fail to import large amounts of data. Python can also obtain C-like speeds when implementing certain software, e.g., Cython, Sage.

The output of Simplex is a broad array of information held in seven .csv files. The most important of these is SimData.csv, and is intended to be analyzed in the freely available R (https://www.r-project.org/) and RStudio (https://www.rstudio.com/) environments. The R statistical computing language is well-suited to the analysis of Simplex output and contains many packages for multivariate analysis and higher-order statistical analysis that Python is only beginning to accumulate. Consequently, I provide R source code in .R files and .Rmd (RMarkdown) files, complete with basic and advanced statistical analyses for analyzing diversity, regression models, ordination, variance partitioning, and for generating pdf documents (via Knitr) that integrate prose, code, and figures for manuscripts.

The reader can view example R-based analysis files that users can use to examine Simplex's simulated data: https://github.com/LennonLab/simplex/tree/master/results/analyses.

@@

\setlength\parindent{36pt}
Simplex models do not complete until the time series of total abundance values reaches a state of mean reversion (i.e., stationarity). Because Simplex models can range from quickly flowing systems with high disturbance to barely flowing and highly stable but depleted systems, simulations can potentially take several minutes or more to complete. Likewise, the ability to simulate many complex scenarios also allows for very large total abundances, the values of which cannot be predicted *a priori* and can even potentially outstrip a computer's memory.

I ran Simplex on a Mid 2010 MacBook Pro (OS X 10.9.5) with a 2.4 GHz Intel Core 2 Duo processor and 4GB of Memory. This system probably represents a below average capacity for modern personal computers, which for this study, was desirable as Simplex should be able to be ran on both personal computers and high capacity remote servers. I present results for time to completion and required memory in the Results.

@@

**Unit tests**  
Simplex passed all units tests for 15 diversity indices, ensuring that each index returns either the correct calculated value, or 'NaN' if given any values that cannot be used (e.g., negative numbers, string characters, empty lists).

**Speed & Memory**  
The following are results from 100 randomly assembled models. On average, Simplex models required 96.60 +- 32.70 seconds to run, had an average total abundance (*N*) of 31,367 +- 879, and required 161 +- 4.3 megabytes of memory. The longest any model took to complete was 49 minutes. This particularly demanding model required 261MB of memory and also had generated the greatest total abundance (*N* = 68,808). The shortest any model took to complete was 5.8 seconds with a total abundance of 0 individuals. The least amount of memory any model required was 90MB. These and other analyses from the 100 randomly assembled models can be run using the SpeedMemory.Rmd file located in the results/TestResults directory.

@@

There are currently two R Markdown analysis files associated with Simplex. These have been included as supplemental files. These are located in the tools/analyses/Rmd directory and are designed to help users understand how to import Simplex files and use R Markdown to analyze Simplex data. The first file, BasicAnalysis.Rmd, focuses on descriptive statistics, multiple univariate analyses, and variance partitioning to reveal relationships among Simplex variables. Running this file on example Simplex data revealed that Simplex produces well-known and/or intuitive relationships (Figs 1, 2, 3, 4).

@@

**Figure 1.** A scatter plot matrix revealing that the values of inputs representing physical variables are independent of one another, as should be expected in the modeling. This is only a subset of the physical variables in Simplex. Amplitude pertains to magnitude of temporal fluctuation in the inflow of propagules and/or resources. Barriers pertains to the number of barriers in the system, which varied between 1 and 4.

@@

**Figure 2.** A scatter plot matrix revealing that the values of resource-related variables. This is only a subset of resource variables in Simplex. Number of resource particles, their concentration, resource richness, and resource diversity (a combination of resource richness and the variance in abundance among resources) are all highly and positively correlated, as should generally be expected.

@@

**Figure 3.** A scatter plot matrix revealing that the values of physiological variables. This is only a subset of those in simplex. As of 29 October 2015, Simplex does not include explicit stoichiometry. An intuitive life history trade-off arises in that increasing growth rate generally leads to increasing metabolic maintenance.

@@

**Figure 4.** A scatter plot matrix revealing that the values of diversity-related variables. This is only a subset of those in Simplex. There are strong positive relationships between total numbers of individuals (total abundance) and total numbers of species (species richness) and between total abundance and the abundance of the most abundant species (Nmax). Each of these relationships is widely observed in ecological communities (Locey and Lennon 2015). Likewise, different measures of species turnover largely reflect one another, as would generally be expected.

@@

**Animations**   
I used the animation module in the matplotlib plotting library to generate avi formatted movies of Simplex models. These can be viewed in any internet brower (https://github.com/LennonLab/simplex/tree/master/results/movies/examples). Currently, generating movies requires following three easy steps explained in model/model.py.
